---
title: "Motor Trend magazine  - Data analysis of influence on MPG for Automatic vs. Manual Transmission."
author: "by jmvilaverde"
date: "Thursday, June 18, 2015"
output: pdf_document
---
## Executive summary

With the linear regression model evaluated have concluded that, with 95% confidence, is estimated that a manual transmission results in a 0.0457303 to 5.8259441 increase in MPG comparing to use an automatic transmission.

***

## 1.Initial Exploratory Data Analysis

```{r dataAdquisition, echo=FALSE}
data(mtcars)
```

### **Structure from ?mtcars and values for factors.** _(View Figure 1.Mtcars pairs comparation.)_

*Format:* _A data frame with 32 observations on 11 variables._

Variables | Units                                         | Values
--------- | --------------------------------------------- | ---------------------------
**mpg**   | **Miles/(US) gallon**                         |
cyl       | Number of cylinders (4,6,8)                   | `r unique(mtcars$cyl)[order(unique(mtcars$cyl))]`
disp      | Displacement (cu.in.)                         |
hp        | Gross horsepower                              |
drat      | Rear axle ratio                               |
wt        | Weight (lb/1000)                              |
qsec      | 1/4 mile time                                 |
vs        | V/S -> V motor or straight motor              | `r unique(mtcars$vs)[order(unique(mtcars$vs))]`
**am**    | **Transmission (0 = automatic, 1 = manual)**  | `r unique(mtcars$am)[order(unique(mtcars$am))]`
gear      | Number of forward gears                       | `r unique(mtcars$gear)[order(unique(mtcars$gear))]`
carb      | Number of carburetors                         | `r unique(mtcars$carb)[order(unique(mtcars$carb))]`

Correlation between mpg and am is `r cor(mtcars$mpg, mtcars$am)`. (Closer to -1 or 1 is stronger relationship, when is 0 implies no linear relationship).

## 2.Model proposal and analysis

### 2.1.Analysis of Model Initial. _(View Figure C1.Figure Summary Model Initial.)_

> $mpg_i={\beta}_0+\beta_{am}am_i$

```{r modelInitial, echo=FALSE}
model.Initial <- lm(mpg ~ am, data=mtcars)
coeff.Model.Initial <- coef(model.Initial)
pvalue.model.Initial <- as.character(format(anova(model.Initial)$'Pr(>F)'[1],scientific=TRUE))
pvalues.coff.model.Initial <- as.character(format(summary(model.Initial)$coefficients[,4],scientific=TRUE))
r.model.Initial <- summary(model.Initial)$r.squared
```

* P-value intercept and coefficients:  `r pvalues.coff.model.Initial` < 0.05 are good p-values.
* P-value Model: `r pvalue.model.Initial`. < 0.05 is a good p-value for the model.
* $R^2$ value: `r r.model.Initial`. **This low value indicates that model Initial only fits 36% of the data.**

#### 2.2.Analysis of model Complete. _(View Figure C2.Figure Summary Model Complete.)_

> $mpg =\beta_{cyl}cyl+\beta_{dips}disp+\beta_{hp}hp+\beta_{drat}drat+\beta_{wt}wt+\beta_{qsec}qsec+\beta_{vs}vs+\beta_{am}am+\beta_{gear}gear+\beta_{carb}carb$

```{r modelComplete, echo=FALSE}
model.Complete <- lm(mpg ~ ., data=mtcars)
coeff.Model.Complete <- coef(model.Complete)
pvalue.model.Complete <- as.character(format(anova(model.Complete)$'Pr(>F)'[1],scientific=TRUE))
pvalues.coff.model.Complete <- as.character(format(summary(model.Complete)$coefficients[,4],scientific=TRUE))
r.model.Complete <- summary(model.Complete)$r.squared
```

* P-value intercept and coefficients:  `r pvalues.coff.model.Complete` > 0.05 are bad p-values.
* P-value Model: `r pvalue.model.Complete`. < 0.05 is a good p-value for the model.
* $R^2$ value: `r r.model.Complete`. This high value indicates that the model Complete is a good fit to the data.

**The p-value for intercept and coefficients indicates that model Complete is not a good model for fit the data.**

#### 2.3.Propose and analysis of alternative model using step R function. _(View Figure C3.Figure Summary Model Step.)_

```{r modelStep, echo=FALSE}
step.result <- as.character(step(model.Complete, trace=FALSE)$call)
model.Step <- lm(mpg ~ wt + qsec + am, data=mtcars)
coeff.Model.Step <- coef(model.Step)
pvalue.model.Step <- as.character(format(anova(model.Step)$'Pr(>F)'[1],scientific=TRUE))
pvalues.coff.model.Step <- as.character(format(summary(model.Step)$coefficients[,4],scientific=TRUE))
r.model.Step <- summary(model.Step)$r.squared
```

> step(lm(mpg ~ ., data=mtcars)): `r step.result`

> $mpg =\beta_{wt}wt+\beta_{qsec}qsec+\beta_{am}am$

* P-value intercept and coefficients:  `r pvalues.coff.model.Step` < 0.05 for coefficients are good p-value for the model.
* p-value Model: `r pvalue.model.Step`. < 0.05 is a good p-value for the model.
* R^2: `r r.model.Step`. This high value indicates that the model Step is a good fit to the data.(1 means 100% fit).

*This model have a p-value < 0.05 for model and coefficients, and a $R^2$ value near 85%. This model is accepted.*

```{r interactionsModelStep, echo=FALSE}
#Confidence Interval
sumCoef.Step <- summary(model.Step)$coef
confInterval.Model.Step <- sumCoef.Step[4,1] + c(-1,1) * qt(0.975, df=model.Step$df) * sumCoef.Step[4,2]
```

_(View Figure 2. Model Step plot.)_

_(View Figure C4. Anova comparation between the models.)_

***

##Final Analysis based on information from Model Step.

### Is an automatic or manual transmission better for MPG?

> The manual transmission mean is higher than automatic transmission mean for Model Step, due to this analysis is determined that manual transmission is better for MPG.

### Quantify the MPG difference between automatic and manual transmissions.

> With 95% confidence, we estimate that a manual transmission results in a `r confInterval.Model.Step[1]` to `r confInterval.Model.Step[2]` increase in MPG comparing to use of automatic transmission for a car with the same weight (wt) and 1/4 mile time (qsec).

***
## Apendix. (Note: Summary and Anova are included as Figures).

```{r pairsMtcars, echo=FALSE, fig.height=5.5, fig.width=7}
pairs(mtcars, panel = panel.smooth, main = "Figure 1. Mtcars pairs comparation.", col=3+(mtcars$am>0))
```
```{r plotModelA, echo=FALSE, fig.height=3, fig.width=7, fig.align='center'}
par(mfrow=c(1,2))
plot(model.Step,col=mtcars$am+1, qqline=TRUE, which=1:2)
title("Figure 2.Model Step Plot.", outer=TRUE, font.main=2, line=-1)
```

```{r plotModelB, echo=FALSE, fig.height=3, fig.width=7, fig.align='center'}
par(mfrow=c(1,2))
plot(model.Step,col=mtcars$am+1, qqline=TRUE, which=3:4)
title("Figure 2.Model Step Plot.", outer=TRUE, font.main=2, line=-1)
```

_Figure C1.Figure Summary Model Initial._
```{r modelInitialPlot, echo=FALSE}
summary(model.Initial)
```

_Figure C2.Figure Summary Model Complete._
```{r modelCompletePlot, echo=FALSE}
summary(model.Complete)
```

_Figure C3.Figure Summary Model Step._
```{r modelStepPlot, echo=FALSE}
summary(model.Step)
```

_Figure C4. Anova comparation between the models._
```{r anova, echo=FALSE}
anova(model.Initial, model.Step, model.Complete)
```
