---
title: "Motor Trend magazine  - Data analysis of influence on MPG for Automatic vs. Manual Transmission."
author: "by jmvilaverde"
date: "Thursday, June 18, 2015"
output: pdf_document
---

##Context

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

“Is an automatic or manual transmission better for MPG”
"Quantify the MPG difference between automatic and manual transmissions"


##Question

Take the mtcars data set and write up an analysis to answer their question using regression models and exploratory data analyses.

Your report must be:

Written as a PDF printout of a compiled (using knitr) R markdown document.
Brief. Roughly the equivalent of 2 pages or less for the main text. Supporting figures in an appendix can be included up to 5 total pages including the 2 for the main report. The appendix can only include figures.
Include a first paragraph executive summary.
Upload your PDF by clicking the Upload button below the text box.

##Peer Grading

The criteria that your classmates will use to evaluate and grade your work are shown below. 
Each criteria is binary: (1 point = criteria met acceptably; 0 points = criteria not met acceptably)
Your Course Project score will be the sum of the points and will count as 40% of your final grade in the course. 

#Evaluation/feedback on the above work

* Did the student interpret the coefficients correctly?
* Did the student do some exploratory data analyses?
* Did the student fit multiple models and detail their strategy for model selection?
* Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
* Did the student do a residual plot and some diagnostics?
* Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
* Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
* Did the report include an executive summary?
* Was the report done in Rmd (knitr)?

##Report brief (about 2 pages)

###Executive summary (First paragraph)

Automatic/Manual transmission is better...


#### Data adquisition

##### 1. Data adquisition and Initial structure analysis

##### 1.1. Variables

```{r dataAdquisition, echo=TRUE}
data(mtcars)
        
#?mtcars

#Show colnames of mtcars
names(mtcars)
```

**mtcars**

Format: _A data frame with 32 observations on 11 variables._

Variable| Units
------- | -----------
**mpg** | **Miles/(US) gallon**
cyl     | Number of cylinders
disp    | Displacement (cu.in.)
hp      | Gross horsepower
drat    | Rear axle ratio
wt      | Weight (lb/1000)
qsec    | 1/4 mile time
vs      | V/S
**am**  | **Transmission (0 = automatic, 1 = manual)**
gear    | Number of forward gears
carb    | Number of carburetors

```{r structureMtcars, echo=TRUE}
#Show structure of mtcars
str(mtcars)
```


##### 2.Initial analysis
Selected variables: 

* Predictor:    X = am,  Transmission with values 0 for automatic, 1 for manual.
* Outcome:      Y = mpg, Miles/(US) gallon

Linear regression model formula:

* Y = ${\beta}_0 + {\beta}_1X$ -> mpg = ${\beta}_0 + {\beta}_1am$

Create initial model:

Calculate $\beta_0$ (Intercept): $\hat{\beta}_0=\bar{Y}+\hat{\beta}_1 \bar{X}$

Calculate $\beta_1$ (Slope): $\hat{\beta}_1= Cor(Y,X)\frac{Sd(Y)}{Sd(X)}$

Fitting the best line:

* $\Sigma_{n}^{i=1}{(Y_i-(\beta_0+\beta_1X_i))}^2$ to minimize the distance.

```{r teoreticalFormula, echo=TRUE}
#Theoretical formula
y <- mtcars$mpg
x <- mtcars$am
beta1 <- cor(y,x)*(sd(y)/sd(x))
beta0 <- mean(y) - beta1 * mean(x)
#R function
model.Initial <- lm(y ~ x)
coeffs <- coef(model.Initial)

#Comparation
rbind(c(beta0, beta1), coeffs)
```

For linear model $Y=\beta_0+\beta_1X$ -> $mpg=\beta_0+\beta_1am$

* Intercept $\beta_0=$ `r beta0`
* Slope $\beta_1=$ `r beta1`

> For this linear regression model, the expected difference in MPG is `r beta1` when the car have manual transmission in comparation to the same car with automatic transmission.
The Intercept `r beta0` is the expected MPG of a automatic transmission car.

```{r plotModelInitial, echo=TRUE}
color <- ifelse(x==0,"red","blue")
plot(x,y,col=color,xlab="x=am",ylab="y=mpg")
legend("top",c("Automatic","Manual"),col=c("red","blue"),pch=1)
abline(model.Initial)
```

_View Figure 1 for plot with regression line._

###Normalizing data????

##### 3.Basic regression model with additive Gaussian errors.

Into point 2 is obtained a linear regression model that fits the data, but doesn't take in consideration the impact of the others variables. We are going to analyze gaussian errors for Initial Model.

Selected variables: 

* **Predictor**:    X = am,  Transmission with values 0 for automatic, 1 for manual.
* **Outcome**:      Y = mpg, Miles/(US) gallon

Probabilistic model for linear regression:

* $Y_i={\beta}_0+{\beta}_1X_i+\epsilon_i$ -> $mpg_i={\beta}_0+{\beta}_1am_i+\epsilon_i$
* $\epsilon_i$ are assumed iid $N(\mu_i,\delta^2)$. 
* Note, $E[Y_i|X_i=x_i]=\mu_i=\beta_0+\beta_1x_i$ and $Var(Y_i|X_i=x_i)=\delta^2$.
* $\hat{\beta}_0=\bar{Y}+\hat{\beta}_1 \bar{X}$ and $\hat{\beta}_1= Cor(Y,X)\frac{Sd(Y)}{Sd(X)}$.

Residuals analysis:

* Observed outcome i is $Y_i$ at a predictor value $X_i$.
* Predicted outcome i is $\hat{Y}_i$ at a predictor value $X_i$ is $\hat{Y}_i=\hat{\beta}_0+\hat{\beta}_1X_i$.
* Residual is the difference between observed an predicted: $e_i=Y_i-\hat{Y}$, the vertical distance between the observed data point and the regression line.
* Least squares minimizes $\Sigma_{i=1}^ne_i^2$.
* $e_i$ can be thought of as estimates of the $\epsilon_i$.
```{r residualsModelInitial, echo=TRUE}
#Calculate residuals
e.ModelInitial <- y-(beta0+beta1*x)
e.ModelInitial
#R function for residuals
resid(model.Initial)

#Calculate predicted y (mpg)
yhat <- predict(model.Initial)
#Calculate max difference between residual and observed Y - predicted Y (y hat)
max(abs(e.ModelInitial -(y - yhat)))
```

###“Is an automatic or manual transmission better for MPG”



###

##Main Body + Apendix only figures (not more than 5)